---
title: "chapter6"
output: html_document
---

# RStudio Exercise 6: Analysis of longitudinal data


Let's load the required addons

```{r echo = T, results = 'hide', message=FALSE}

library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(lme4)
library(lmerTest)

```
The BPRS data contains data from 40 males, equally assigned to two treatment groups(20 each group). Brief psychiatric scale rating was measured at baseline i.e. before treatment and then every week subsequently, until week 8. BPRS is used to evaluate patients that are suspected of having schizophrenia.

The RATS data is from a nutrition study carried out on RATS. Each group was put on a different diet and their weight was followed for a period of 9 weeks. Measurements were taken every week, except week six when two measurements were taken. 

The data in these datasets has been converted into long form. The script for creating long form data can be found in the data folder and the difference between both is explained in the script file.

Let's read the datasets BPRS and RATS created using the instructions form the data wrangling exercise.

```{r echo = TRUE}

BPRS <- read.csv("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep=" ")
RATS <- read.csv("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", sep='\t')


RATSL <- read.table("data/RATSL.csv", sep=",", header=TRUE, check.names=FALSE, row.names = 1)

BPRSL <- read.table("data/BPRSL.csv", sep=",", header=TRUE, check.names=FALSE, row.names = 1)


glimpse(RATSL)
glimpse(BPRSL)


```



As the datasets were saved in csv format, after reading, the factor variables have turned into integers. Let's convert them back to factor variables. This is important, so that the software knows that these are categorical and not a scalar variables.


```{r echo = TRUE}

BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)



glimpse(RATSL)
glimpse(BPRSL)

```


### Chapter 8 analyses on RATS dataset

Now we will be working with RATS dataset and as instructed, we will carry out the analyses in chapter 8 of Kimmo's textbook, but using the RATS dataset instead of BPRS.

Let's visualize the data.

```{r echo = TRUE}

ggplot(RATSL, aes(x = RATSL$Time, y = RATSL$Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))




```

Here we can see can already see that within the three groups of rats, there is a difference in baseline weight. We can also see that there is an upward trend for weight in all groups.

What is apparent is that the rats with higher weight continued to have high weight throughout the study. This phenomenon is called tracking and we shall try to visualize it by **"subtracting the relevant occasion mean from the original observation and then dividing by the corresponding visit standard deviation."** (source, Kimmo's book)



```{r echo = TRUE}
# Standardizing weight
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdweight = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()



# Plotting again with standardized weight

ggplot(RATSL, aes(x = Time, y = stdweight, linetype = ID)) +
  geom_line() +
  theme(legend.position = "none") + 
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = RATSL$stdweight)

```

The large number of individuals here leave a lot to the imagination though. It would be better is we could draw a more concise plot showing mean profiles for each group, and the variation between these measurements. 


```{r echo = TRUE}

n <- RATSL$Time %>% unique() %>% length()

RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = (sd(Weight) /  
   sqrt(n))) %>%
  ungroup()

# Plotting the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.5)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")


```
Another way to visualize data is through boxplots.

```{r echo = TRUE}
RATSL8S <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weigth), Time 8-64")


```
We can atleast identify one outlier that is affecting the results. Let's remove it and redraw the plot.

```{r echo = TRUE}
RATSL8S1 <- RATSL8S%>% filter(mean < 590 )



ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=2, fill = "white") +
  scale_y_continuous(name = "mean(Weigth), Time 8-64")

```



Now it is time to run the t test. However, the problem is that t test is meant for comparison between two group, but here we have three groups. We will instead use One-way ANOVA for within group comparison. 


```{r echo = TRUE}

RATSL8S1owa <- aov(mean ~ Group, data = RATSL8S1)

summary(RATSL8S1owa)
```
The results from ANOVA tell us that the groups differ significantly between each other based on weight gain. However to find out which groups are different we need to run a post-hoc analysis.

```{r echo = TRUE}

TukeyHSD(RATSL8S1owa)
```

Results from Tukey post-HOC test show that it is group 1 that differs from group 2 and 3 significantly. 

But the problem is that we have not adjusted the model for baseline weight. That can be an important factor considering that the rats in group a had a significantly lower weight at the start of the study (as seen in the plots).

Let's run a linear regression modal with baseline values as of the predictors.

```{r echo = TRUE}
RATS
# Creating a new variable 
RATSL8S2 <- RATSL8S %>%
  mutate(baseline = RATS$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = RATSL8S2)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)

```

After adjusting for baseline values, the group differences are not significant anymore!
Thus these nutritional interventions did not cause a difference in weight gain in any of these groups.

### Chapter 9 analyses on BPRS dataset


In this section we will run linear mixed effects models for analysis. These models use random effects (within group and within individual variation in repeated measures) as predictor variables in the models. Repeated measures are correlated with each other, BUT When adjusted for these random effects, the repeated measured are considered independent.

First, let's try to visualize the dataset in its long form. The graph shows BPRS scores for individuals in both treatment groups. I have drawn a loess curve in order to show the trend for the whole group. Also let's make a matrix of BPRS scores.



```{r echo = TRUE}

ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +geom_line() + stat_summary(aes(group = 1), geom = "point", fun.y = mean, shape = 17, size = 3) + facet_grid(. ~ treatment) + scale_x_continuous(name = "Time (weeks)") +scale_y_continuous(name = "BPRS") + theme(legend.position = "none") + stat_smooth(aes(group = 1)) 

pairs(BPRS[3:11],  col = BPRS$treatment)
```

Let's run a simple linear regression that assumes that these repeated measures are independent

```{r echo = TRUE}
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)


summary(BPRS_reg)
```
A simple regression model tells us that there are no significant differences between groups based on treatment. 

Let's now run a random intercept model which identifies the repeated nature of measures and also allows the linear regression fit for each individual rat separately. 



```{r echo = TRUE}

BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

summary(BPRS_ref)

```

Note that we have also included random effects that originate from within individual variation. 

Now lets add time (week) into the random effects.


```{r echo = TRUE}

BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

summary(BPRS_ref1)

```


```{r echo = TRUE}

anova(BPRS_ref1, BPRS_ref)

```


Compared to the previous model, the new model that includes both week and subject random effects is significantly better.

Finally let's also add the interaction between week and treatment group variables (i.e. fixed effect). 

```{r echo = TRUE}

BPRS_ref2 <- lmer(bprs ~ week + treatment + (week * treatment)+ (week|subject), data = BPRSL, REML = FALSE)

summary(BPRS_ref2)
```

```{r echo = TRUE}
anova(BPRS_ref2, BPRS_ref1)
```

This does not result in a significant improvement in our new model, which demonstrates that there is little or no interaction between treatment group and week variables.

In this model and all previous models, there was no significant difference in BPRS between treatment groups.


Let's now draw a new plot based on our fitted model and compare it with the plot we drew at the beginning. 



```{r echo = TRUE}
ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +geom_line() + stat_summary(aes(group = 1), geom = "point", fun.y = mean, shape = 17, size = 3) + facet_grid(. ~ treatment) + scale_x_continuous(name = "Time (weeks)") +scale_y_continuous(name = "BPRS") + theme(legend.position = "none") + stat_smooth(aes(group = 1)) 

```


```{r echo = TRUE}


Fitted <- fitted(BPRS_ref2)


BPRSL <- mutate(BPRSL, Fitted = Fitted)

ggplot(BPRSL, aes(x = week, y = Fitted, group = subject)) +geom_line() + stat_summary(aes(group = 1), geom = "point", fun.y = mean, shape = 17, size = 3) + facet_grid(. ~ treatment) + scale_x_continuous(name = "Time (weeks)") +scale_y_continuous(name = "BPRS") + theme(legend.position = "none") + stat_smooth(aes(group = 1)) 

```

Notice how the model fits regression lines for all individuals and also notice the final regression line is fitted. One can visually appreciate as well that there is no difference in BPRS score trend between groups.

